\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move

### 2025-09-02 writing/appendix-b-gams.Rmd
moved model code to appendix B

@@

```{r kept-hrsf}
readr::read_csv('../data/tracking-data/telemetry-metadata.csv',
                show_col_types = FALSE) %>%
  mutate(keep_for_RSF = if_else(keep_for_RSF == 'yes', 'yes', 'no')) %>%
  group_by(species, keep_for_RSF) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  # fix species names
  mutate(species = case_when(
    species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (sm)',
    species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (b)',
    .default = species)) %>%
  mutate(species = paste0('\\emph{', species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .)) %>%
  bind_rows(.,
            group_by(., keep_for_RSF) %>%
              summarize(species = 'Total',
                        n = sum(n)) %>%
              mutate(prop = n / sum(n))) %>%
  rename(Species = species, `Kept` = keep_for_RSF, Proportion = prop) %>%
  knitr::kable(format = 'markdown', caption = "Number of animals (n) and proportion of animals (Proportion) used or excluded (Kept) when fitting the hierarchical resource selection function. The models only used animals that showed stationary range residency or two clear seasonal ranges within the tracking period. We excluded animals with insufficient data to estimate a range size and animals that migrated but did not establish clear ranges within the tracking period.")
```

>>>>

\noindent The two models for the probability of moving and speed when moving were fit using the code below:

```{r hgam_1, eval=FALSE, echo=TRUE}
m_moving <- bam(moving ~ # model estimating probability of moving
                  # random intercept for each animal
                  s(animal, bs = 're') +
                  # fixed intercept for each species
                  species +
                  # to account for changes in behavior within days
                  s(tod_pdt, by = species, k = 5, bs = 'cc') +
                  # to account for changes in behavior within years
                  s(doy, by = species, k = 5, bs = 'cc') +
                  # species-level effect of temperature
                  s(temp_c, by = species, k = 5, bs = 'tp') +
                  # to account for seasonal changes in day length
                  ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
                  # to account for changes in day nocturnality with temperature
                  ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
                  # to account for changes in fur coats seasonally
                  ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
                  # larger sampling intervals underestimate movement speed
                  s(log(dt), k = 3) +
                  s(log(dt), species, k = 3, bs = 'fs'),
                family = binomial(link = 'logit'),
                data = d,
                method = 'fREML', # fast REML
                discrete = TRUE, # discretize the covariates for faster computation
                knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5)), #' for `bs = 'cc'`
                control = gam.control(trace = TRUE))
```

```{r hgam_2, eval=FALSE, echo=TRUE}
m_speed <- bam(speed_est ~ # model estimating speed when moving (i.e., speed > 0)
                 # random intercept for each animal
                 s(animal, bs = 're') +
                 # fixed intercept for each species
                 species +
                 # to account for changes in behavior within days
                 s(tod_pdt, by = species, k = 5, bs = 'cc') +
                 # to account for changes in behavior within years
                 s(doy, by = species, k = 5, bs = 'cc') +
                 # species-level effect of temperature
                 s(temp_c, by = species, k = 5, bs = 'tp') +
                 # to account for seasonal changes in day length
                 ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
                 # to account for changes in day nocturnality with temperature
                 ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
                 # to account for changes in fur coats seasonally
                 ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
                 # larger sampling intervals underestimate movement speed
                 s(log(dt), k = 3) +
                 s(log(dt), species, k = 3, bs = 'fs'),
               family = Gamma(link = 'log'), # can use Gamma because no zeros
               data = d_2,
               method = 'fREML', # fast REML
               discrete = TRUE, # discretize the covariates for faster computation
               knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5)), #' for `bs = 'cc'`
               control = gam.control(trace = TRUE))
```

@@

\noindent Each species had a Hierarchical Resource Selection Function fit with the code below. See @aarts_estimating_2008 for additional information on fitting Resource Selection Functions using Poisson models.

```{r rsf, eval=FALSE, echo=TRUE}
rsf <- bam(detected ~ # 1 for telemetry locations, 0 for quadrature points
             # species-level average resource preference
             s(forest_perc, k = 4, bs = 'cr') +
             s(elevation_m, k = 4, bs = 'cr') +
             s(dist_water_m, k = 4, bs = 'cr') +
             # animal-level deviations from the species-level average
             s(animal, bs = 're') +
             s(forest_perc, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
             s(elevation_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
             s(dist_water_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
             # changes in preference with temperature
             ti(forest_perc, temperature_C, k = 4, bs = 'cr') +
             ti(elevation_m, temperature_C, k = 4, bs = 'cr') +
             ti(dist_water_m, temperature_C, k = 4, bs = 'cr') +
             # include marginals of temperature to post-stratify over afterwards
             s(temperature_C, k = 4, bs = 'cr') +
             s(temperature_C, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
           family = poisson(link = 'log'),
           data = d,
           weights = weight, # based on AKDE
           subset = species == sp, # subset to the species of interest
           method = 'fREML',
           discrete = TRUE)
```

```{r kept-hrsf}
readr::read_csv('../data/tracking-data/telemetry-metadata.csv',
                show_col_types = FALSE) %>%
  mutate(keep_for_RSF = if_else(keep_for_RSF == 'yes', 'yes', 'no')) %>%
  group_by(species, keep_for_RSF) %>%
  summarize(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  # fix species names
  mutate(species = case_when(
    species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (sm)',
    species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (b)',
    .default = species)) %>%
  mutate(species = paste0('\\emph{', species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .)) %>%
  bind_rows(.,
            group_by(., keep_for_RSF) %>%
              summarize(species = 'Total',
                        n = sum(n)) %>%
              mutate(prop = n / sum(n))) %>%
  rename(Species = species, `Kept` = keep_for_RSF, Proportion = prop) %>%
  knitr::kable(format = 'markdown', digits = 3, caption = "Number of animals (n) and proportion of animals (Proportion) used or excluded (Kept) when fitting the hierarchical resource selection function. The models only used animals that showed stationary range residency or two clear seasonal ranges within the tracking period. We excluded animals with insufficient data to estimate a range size and animals that migrated but did not establish clear ranges within the tracking period.")
```
